PAPER DRAFT

### 2021-11-10 Papers/Polymethylation_scores/02_abstract.Rmd

**Background**: Prenatal maternal smoking is a difficult to measure prenatal exposure with negative implications for child health.  DNA methylation signatures in blood and placental samples can function as a biomarker of prenatal maternal smoking. However little work has assessed how DNA methylation signatures of prenatal maternal smoking vary across ages, ancestry groups or tissues. Using children in the Fragile Families and Child Wellbeing study, we tested if prenatal maternal smoking is associated with salivary polymethylation scores. We assessed the consistency of associations at ages 9 and 15 and their portability across  African, European and Hispanic ancestries.

**Results**: We created a polymethylation score using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood. At age 9, prenatal maternal smoke exposure was associated with a 0.52 (95%CI: 0.36, 0.67) standard deviation higher polymethylation score. At age 15, At age 9, prenatal maternal smoke exposure was associated with a 0.46 (95%CI: 0.3, 0.62) higher polymetylation score.  The direction and magnitude of the association was portable across genetic ancestries at age 9 although the association was attenuated among European and Hispanic genetic ancestry samples at age 15. The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (Area under the curve from receiver operating curve (AUC-ROC age 9: 0.77 age 15: 0.77), and more accurate than using a single *a priori* CpG site, AHRR:CpG05575921 (AUC-ROC age 9:0.74, *P* value=0.02; age 15:0.73, *P* value=0.02).

**Conclusions**: Prenatal maternal smoking associates with DNA methylation signatures in saliva samples, a clinically practical tissue. Polymethylation scores for prenatal maternal smoking are portable across ancestries and more accurate than some individual CpG sites.  Thus, DNA polymethylation scores from saliva samples could serve as robust and practical clinical biomarkers of prenatal maternal smoke exposure.

>>>>

**Background**: Prenatal maternal smoking has negative implications for child health.  DNA methylation signatures  can function as a biomarker of prenatal maternal smoking. However little work has assessed how DNA methylation signatures of prenatal maternal smoking vary across ages, ancestry groups, or tissues. In the Fragile Families and Child Wellbeing study, we tested whether prenatal maternal smoking was associated with salivary polymethylation scores. We assessed the consistency of associations at ages 9 and 15, their portability across  participants from African, European, and Hispanic genetic ancestries and the accuracy of exposure classification using Area Under the Curve (AUC) from receiver operating curve analyses.

**Results**: We created saliva polymethylation scores using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood. In the full sample at age 9 (n=756), prenatal maternal smoke exposure was associated with a 0.52 (95%CI: 0.36, 0.67) standard deviation higher polymethylation score. The direction and magnitude of the association was consistent across when stratified genetic ancestries. In the full sample at age 15 (n=749), prenatal maternal smoke exposure was associated with a 0.46 (95%CI: 0.3, 0.62) standard deviation higher polymetylation score, and the association was attenuated among the European and Hispanic genetic ancestry samples. The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (AUC age 9=0.77, *P* value<0.001, age 15=0.77, *P* value<0.001), and more accurate than using a single *a priori* site in the AHRR gene (cg05575921 AUC=0.74, *P* value=0.03; age 15=0.73, *P* value=0.01).

**Conclusions**: Prenatal maternal smoking was associated with DNA methylation signatures in saliva samples, a clinically practical tissue. Polymethylation scores for prenatal maternal smoking were portable across genetic ancestries and more accurate than individual DNA methylation sites. DNA polymethylation scores from saliva samples could serve as robust and practical clinical biomarkers of prenatal maternal smoke exposure.

### 2021-11-10 Papers/Polymethylation_scores/04_background.Rmd

Maternal cigarette smoking during pregnancy is associated with reproducible DNA methylation signatures, measured in infant cord blood and placental tissues, and primarily tested in European cohorts [@Joubert:2016ue; @Everson:2021td; @Cardenas2019]. Prenatal maternal cigarette smoking exposure is associated with postnatal DNA methylation signatures in blood from cohorts of older children and adults[@Tehranifar:2018tb ; @Richmond:2018tu]. Few studies have considered the persistence of the association between maternal prenatal cigarette smoking and DNA methylation at different ages in the same individuals. Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation across tissues and ancestry groups. DNA methylation drives cell differentiation and cell type proportions differ across tissue types. Epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples [@Everson:2021td]. Cord blood and placental samples are unlikely to be available in clinical settings. While peripheral blood is a more realistic clinical samples, saliva is even easier to collect. To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation. Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017], and DNA methylation signatures of own-smoking differ across ancestry groups [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012]. Only a few studies have examined associations between prenatal maternal cigarette smoking and DNA methylation in non-European ancestry populations. Among 954 infants, 70% of whom had mothers identifying as Black, prenatal maternal smoking was associated with cord blood DNA methylation at 38 CpG sites [@Xu2021]. The direction of these association between prenatal maternal cigarette smoking and DNA methylation at these sites were generally similar in children of Black and non-Black participants [@Xu2021]. Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking was associated with DNA methylation in peripheral blood samples but ancestry-stratified analyses were not performed [@Ladd-Acosta:2016aa]. Similarly, among 89 middle-aged women, of whom reported African American or Hispanic ethnicity, prenatal maternal cigarette smoking was associated with DNA methylation at 17 of 190 tested CpG sites [@Tehranifar:2018tb]. Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb]. Of 148 CpG sites selected based on their association prenatal maternal smoking among primarily European children, 7 CpG sites were associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue]. Thus, while prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across age, tissue and ancestry is understudied.

An important next step after association testing is the evaluation of DNA methylation as a biomarker of prenatal maternal smoke exposure, which is often done by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior. <!--#DEFINE POLYMETHYLATION SCORES HERE?-->DNA methylation signatures from cord blood classified maternal smoking behavior with accuracy of 91% in a primarily European cohort [@Reese:2017ts]. Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking behavior could also be classified with an accuracy of 81% when using a support vector machine approach and DNA methylation at 26 sites measured in peripheral blood at a single time point [@Ladd-Acosta:2016aa]. Among middle-aged adults, peripheral blood from a single time point could even predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu]. However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across tissue, age and ancestry has not been evaluated. Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times or evaluated if the inclusion of multiple time points improves performance. No study has evaluated the accuracy of salivary DNA methylation as a biomarker for prenatal maternal smoking.

DNA methylation changes hold promise as a biomarker for prenatal maternal smoking. However, before translation to clinical applications, we must understand how and if the signal varies across age, ancestry and tissue sample. In the Fragile Families and Child Wellbeing study, a diverse longitudinal cohort of children, we aimed to assess the potential of salivary DNA methylation biomarkers for prenatal maternal cigarette smoking. We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites. We then performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a persistent and portable biomarker of prenatal maternal smoking.

>>>>

Maternal cigarette smoking during pregnancy is associated with reproducible DNA methylation signatures, measured in infant cord blood and placental tissues, and primarily tested in European cohorts [@Joubert:2016ue; @Everson:2021td; @Cardenas2019]. Prenatal maternal cigarette smoking exposure is associated with postnatal DNA methylation signatures in blood from cohorts of older children and adults [@Tehranifar:2018tb ; @Richmond:2018tu]. Few studies have considered the persistence of the association between maternal prenatal cigarette smoking and DNA methylation at different ages in the same individuals. Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation across tissues. DNA methylation drives cell differentiation and cell type proportions differ across tissue types. Epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples [@Everson:2021td]. Cord blood and placental samples are unlikely to be available in clinical settings. While peripheral blood is a more realistic clinical sample, saliva is even easier to collect. To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation.

Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017], and DNA methylation signatures of own-smoking differ across genetic ancestry groups [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012]. Only a few studies have examined associations between prenatal maternal cigarette smoking and DNA methylation in non-European ancestry populations. Among 954 infants, 70% of whom had mothers identifying as Black, prenatal maternal smoking was associated with cord blood DNA methylation at 38 CpG sites [@Xu2021]. The direction of the associations between prenatal maternal cigarette smoking and DNA methylation at these sites were generally similar in children of Black and non-Black participants [@Xu2021]. Among 572 3-5 year old children, 186 of whom were of African or admixed genetic ancestry, prenatal maternal smoking was associated with DNA methylation in peripheral blood samples but ancestry-stratified analyses were not performed [@Ladd-Acosta:2016aa]. Similarly, among 89 middle-aged women, of whom reported African American or Hispanic ethnicity, prenatal maternal cigarette smoking was associated with DNA methylation at 17 of 190 tested CpG sites [@Tehranifar:2018tb]. Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb]. Of 148 CpG sites selected based on their association with prenatal maternal smoking among primarily European children, 7 CpG sites were also associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue]. While prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across genetic ancestry is understudied.

After association testing, an important next step is to evaluate DNA methylation as a biomarker of prenatal maternal smoke exposure by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior. Biomarkers can consist of single DNA methylation sites or summary meausres across multiple sites [@Philibert2020; @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa]. One attractive option is a polymethylation score, analogous to a polygenic score. In the polymethylation score, CpG sites are weighted by the strength of their association with maternal smoking from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a]. One such DNA methylation score calculated from cord blood classified maternal smoking behavior with an area under the curve (AUC) of 0.82 in a primarily European cohort [@Reese:2017ts]. Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking behavior was classified with an AUC of 0.87 [@Ladd-Acosta:2016aa]. Among middle-aged adults, a score calculated using DNA methylation in peripheral blood samples from a single time point could even predict prenatal smoke exposure (30 years previously) with an AUC of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu]. Ideally, a biomarker for prenatal maternal smoking would be consistent over participant age at DNA methylation sampling and portable across genetic ancestry groups. However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across tissue, age, and ancestry has not been evaluated in the same study.

DNA methylation differences hold promise as a biomarker for prenatal maternal smoking. Before translation to clinical applications, we must understand how and if the signal varies across age, ancestry, and tissue sample. In the Fragile Families and Child Wellbeing study, a diverse longitudinal cohort of children, we aimed to assess the potential of salivary DNA methylation biomarkers for prenatal maternal cigarette smoking. We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites. We then performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a persistent and consistent biomarker of prenatal maternal smoking.
